### 2024-08-28 manuscripts/05-Hussain-Hadah/Hadah-Hispanic-Last-Names.tex
....

\affiliation[]{organization={Department of Economics, Tulane University},
            addressline={6823 St. Charles Ave.},
            city={New Orleans},
%          citysep={}, % Uncomment if no comma needed between city and postcode
            postcode={70118},
            state={LA},
            country={United States}}

\begin{abstract}
\singlespacing % Set single spacing for the abstract
In this paper, I study discrimination against Hispanics in the labor market. I compare the children of inter-ethnic marriages to study the impact of having a Hispanic last name. While males born to Hispanic father-White mothers earn 5 percentage points less than those born to White father-Hispanic mothers, the gap could be completely explained by educational differences. I also study the effect of identifying as Hispanic on earnings. I find that men with a Spanish-sounding last name who identify as Hispanic earn significantly less than those who do not identify as Hispanic, but the gap could also be explained by educational differences.
\end{abstract}

@@

\maketitle
\section{Introduction}\label{sec:intro}

A large literature documents substantial earnings gaps across race and ethnicity \citep{bayer2018divergent, charles2008prejudice, card1992school, fryer2004causes, rubinstein2014pride, bertrand2004emily, juhn1991accounting}. Hispanics constitute a large and growing portion of the population in the United States, see figure \ref{fig:3}. As the number of Hispanics increases, determining whether ethnic discrimination affects their labor market outcomes becomes increasingly crucial \citep{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017,abramitzkyImmigrantsAssimilateMore2020a, abramitzkyNationImmigrantsAssimilation2014,abramitzkyCulturalAssimilationAge2016,chettyWhereLandOpportunity2014}. Thus, it is important to understand whether a person's ethnicity affects their labor market outcomes. Assimilation and mobility are crucial because they reflect how well Hispanics can integrate into society and move up the socioeconomic ladder.

In this paper, I answer the following questions. Does having a Hispanic last name affect labor market outcomes? What is the effect of identifying as Hispanic on earnings? Moreover, I aim to show that comparing Hispanic Whites to non-Hispanic Whites might create an artificially higher earnings gap since the two groups differ on many observable characteristics.\footnote{By identifying as Hispanic, I refer to individuals who self-report their ethnicity as Hispanic on surveys or other data collection instruments.} \footnote{Observable characteristics refer to factors that can be measured and quantified, such as education level, work experience, and immigration status.} Others have attempted to compare how native-born White Hispanics fare to non-Hispanic Whites and foreign-born Hispanics. In \citet{antman2020ethnic,antmanEthnicAttritionObserved2016,antmanEthnicAttritionObserved2016a,antmanEthnicAttritionAssimilation2020}, the authors compare the health and educational outcomes of Hispanic Whites to non-Hispanic Whites and native-born Hispanics to foreign-born Hispanics. They find gaps in education and health between Hispanics and Whites. They also find that native-born Hispanics are more likely than their foreign-born counterparts to report poor health. \citet{davilaChangesRelativeEarnings2008} documents many gaps in labor market outcomes between Hispanics and Whites. They attribute a big part of this gap to differences in education, experience, immigration status, and regional differences.

The US population is growing in diversity. The proportion of non-Whites has increased by more than 10 percentage points from 13 percent in 1995 to 23 percent in 2019. The number of Hispanics has grown by 9 percentage points from 9 percent in 1995 to 16 percent in 2019.\footnote{The portion of non-Whites and Hispanics is calculated using the Current Population Survey (CPS).} Native-born White Hispanic men earn 21\% less than White men, although a substantial portion of the earnings gap is due to educational differences between Hispanics and Whites \citep{duncan2006hispanics, duncan2018identifying, duncan2018socioeconomic}. Some of the earnings differences may also be due to discrimination which will have negative consequences. For example, discrimination against Hispanics can lead to reduced job opportunities, lower wages, and hinder assimilation. In this paper, I examine the role of having a Hispanic last name and identifying as Hispanic on labor market outcomes.

Identifying discrimination is difficult because of factors that affect labor market outcomes that are unobservable to economists---such as unobserved skills. One strategy used by researchers is audit or resume studies. \citet{bertrand2004emily} conducted an audit study where identical resumes were sent to employers with White and Black-sounding names. This approach, however, has its drawbacks. Audit studies only observe callbacks, not wages.

\begin{figure}[t]
\centering
\includegraphics[scale=0.50]{Fig/Hispanics_all.png}
\captionsetup{width=0.85\linewidth} % Adjust the width to your preference
\caption{Hispanics as a percent of the population in the United States from 1994 to 2021.}
\label{fig:3}

\begin{minipage}{\linewidth}
\footnotesize
\textit{Source:} The data source is the Current Population Survey (CPS) from 1994 to 2021.\par
\textit{Note:} The graph shows the percent of the population in the US that self-identifies as Hispanic.
\end{minipage}
\end{figure}

This study utilizes a method developed by \citet{rubinstein2014pride}. I compare children from inter-ethnic marriages. More precisely, I compare children of Hispanic fathers and White mothers (henceforth HW) to children of White fathers and Hispanic mothers (henceforth  WH). This approach stems from the fact that marriages are not random. Couples match on several observable characteristics like income, schooling, socio-economic background, etc. \citep{averettBetterWorseRelationship2008, averettEconomicRealityBeauty1996, beckerTheoryMarriagePart1973, beckerTheoryMarriagePart1974, beckerTreatiseFamily1993, browningCollectiveUnitaryModels2006, chiapporiFatterAttractionAnthropometric2012}. Children of HW and WH marriages have more similar observable characteristics than children of endogamous/homogamous marriages--- i.e., White fathers-White mothers and Hispanic fathers-Hispanic mothers. Moreover, children from a Hispanic father and White mother household will have a Hispanic last name from their fathers, enabling the investigation of how ethnic signals, such as having a Hispanic last name, affect annual log earnings.

Using the Current Population Survey (CPS) from 1994 to 2019 and the 1960 to 2000 US censuses, I estimate the effect of having a Hispanic last name on labor market outcomes. In other words, I approximate the variation in labor market perception of ethnic signals, in this case having a Hispanic last name, on annual log earnings. I also study the effect of identifying as Hispanic on labor market outcomes.

I find that while men born to Hispanic father-White mothers earn less than men born to White father-Hispanic mothers, this gap is entirely explained by educational differences. Thus, I do not find a significant effect of having a Hispanic last name. Finally, I find that men identifying as Hispanic earn significantly less than those who do not, even when I control for ancestry and education.

Previous studies have also used names as a proxy for race and ethnicity \citep{fryer2004causes, rubinstein2014pride, bertrand2004emily}. \citet{fryer2004causes} point out that names can be a predictor of a person's race. Specifically, they provide a rising pattern among Blacks having different names than Whites. They did, however, find that having a Black name, after controlling for the home environment at birth, does not affect their labor market outcomes. \citet{rubinstein2014pride} compared the children of mixed marriages between Sephardic and Ashkenazi Jews in Israel.\footnote{Ashkenazi and Sephardic are two distinct Jewish ethnic groups.} They found that workers with Sephardic last names earn substantially less than those with Ashkenazi last names. \citet{bertrand2004emily} conducted an audit study by sending employers identical resumes that differ in the ethnic and racial signal of a name (Black sounding name versus a White sounding one). They found that resumes with Black-sounding names received substantially fewer callbacks than their White counterparts.

The rest of the paper is organized as follows. Section \ref{sec:data} provides an overview of the data and summary statistics of the sample. Section \ref{sec:emp_model} introduces the empirical framework. I report the results in Section \ref{sec:results}. Finally, I offer a brief conclusion in Section \ref{sec:con1}.

@@

I use two datasets:  the Integrated Public Use Microdata Series (IPUMS) Current Population Survey (CPS) Annual Social and Economic (ASEC) \citep{cps2019} and the 1960 to 2000 US censuses \citep{acs2019}.

I use the CPS data set to study the effect of having a Hispanic last name on a person's labor market outcomes. I take advantage of the fact that the CPS asks parents' place of birth, ethnicity, and race. The data spans the period between 1994--- the earliest sample to ask about a parent's place of birth--- to 2019. Moreover, since the CPS does not provide data on parents' characteristics, essential to determine the family background, I use the census to construct synthetic parents. The census offers a larger sample of potential parents. Similar to the CPS, the 1960 to 2000 censuses ask about the person's place of birth and the individual's race and ethnicity. I employ this information to construct "synthetic parents" using a method developed by \citet{rubinstein2014pride}. I construct the synthetic parents by linking husbands and wives in the census data to each other. I assume that parents have children between the ages of 25 and 40. I then link these synthetic parents using the birth year of child, and place of birth of parents to the year of birth of the "children" and the palce of birth of parents in the CPS sample.

@@

I use the CPS for my primary analysis of the effect of having a Hispanic last name on earnings. I restrict my sample to Whites, United States-born citizens aged 25 to 40 and born between 1960 to 2000. Taking advantage of data on parents' place of birth, I divide the sample into four groups depending on their parent's ethnicity. Mothers or fathers are Hispanic if they were born in a Spanish-speaking country and Puerto Rico, and White if they were born in the United States.\footnote{The list of Spanish Speaking countries include: Argentina, Bolivia, Chile, Colombia, Costa Rica, Cuba, Dominican Republic, Ecuador, El Salvador, Guatemala, Equatorial Guinea, Honduras, Mexico, Nicaragua, Panama, Paraguay, Peru, Spain, Uruguay, Venezuela.} Therefore, an observation can be the product of four types of parents:
\begin{enumerate}
\item White father and White mother (hereafter WW)
\item White father and Hispanic mother (hereafter WH)
\item Hispanic father and White mother (hereafter HW)
\item Hispanic father and Hispanic mother (hereafter HH).
\end{enumerate}

\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{Fig/PirChart2.png}
\caption{Distribution of the four types of children.}
\label{fig:dist}
\begin{minipage}{\linewidth}
\footnotesize
\textit{Source:} The data source is the Current Population Survey (CPS) from 1994 to 2021.\par
\end{minipage}
\end{figure}

\begin{figure}[t]
\centering
\begin{tikzpicture}[node distance =2cm]
\node (start) [startstop] {Synthetic Parents: Observed in 1960 to 2000. $25 \leq age \leq 40$, White, with kids};

\node (pro1) [process, below of = start, yshift = -2cm] {Children: Observed between 1994 and 2019. Born between 1954 to 1999 and $25 \leq age \leq 40$};

\draw [arrow] (start) -- (pro1);
\end{tikzpicture}
\caption{Chart explaining which synthetic parents and children and when we observe them.}
\label{flowchart1}
\end{figure}

The distribution of the four types of children is presented in figure \ref{fig:dist} and table \ref{tab:mat1}. The majority of the sample (95.96\%) is WW children. The second biggest group is HH, which constitutes 2.69\% of the sample. Inter-ethnic children, WH and HW, make up 1.35\% of the sample with 90,325 observations. Even though WH and HW are only 1.35\% of the sample, I have plenty of observations to carry out an analysis. The summary statistics for the children of the four types of marriages are presented in table \ref{tab:c&p1}. Children of WW marriages (column 1) do better on every measure while children of HH parents (column 4) do worse than other children on every measure.

%table1
\input{tables/tab01-observations-by-parents}
%table2
\input{tables/tab02-observations-by-couple}

On the one hand, WW children are more educated, with approximately 14 years of education for men and women. WW children have an unemployment rate of 4\% for men and women. They have log hourly earnings of 2.51 for men and 2.32 for women. Also, their annual log earnings are 10.29 for men and 9.43 for women.

On the other hand, children of HH marriages have an average of 12.9 and 13.24 years of education for men and women, respectively. HH men and women have an unemployment rate of 7\% and 6\%. Their annual log earnings are 10.01 and 9.53 for men and women, respectively.

Inter-ethnic children, WH and HW, have lower educational attainment and income than WW children, but they are better than HH children. Also, WH and HW are comparable, unlike WW and HH children. These two groups are important to the analysis. Children of HW are going to be the treatment group. HW children have a Hispanic father with a Hispanic last names. At the same time, the WH children have White fathers with a non-Hispanic, or White, last names. Summary statistics of these two groups, table \ref{tab:c&p1} columns (2) and (3) can show that they are different from the children of non-inter-ethnic marriages.

%table3
\input{tables/tab03b-summary-stats}

WH children have an average of 13.58 and 13.79 years of education for men and women. The unemployment rate for men and women is 5\%. HW men and women obtained 13.22 and 13.42 years of schooling. Men's unemployment rate is 7\% and women's 6\%.

In panel C of table \ref{tab:c&p1}, I present the rates at which members of each group self-identify as Hispanic. \footnote{Hispanic self-identification is observed when an individual notes that they are Hispanic on the CPS questionnaire} The vast majority of HH children, 96\% for men and 97\% for women, identify as Hispanic. There is considerable attrition in Hispanic identification of inter-ethnic children. Among WH children, 74\% of men and 78\% of women identify as Hispanic. Among HW children, 83\% of men and 82\% of women identify as Hispanic. These results align with the findings of a substantial body of literature that noted this attrition among children of immigrants, especially those coming from inter-ethnic marriages \citep{duncan2017complexity, duncan2018identifying, duncan2020new, antman2020ethnic}. Finally, a small portion of WW children, 4\% of men and 5\% of women, identified as Hispanic. These 5\% of WW children are most likely third generation--- or higher--- children of immigrants. For my analysis, I dropped the WW observations who identify as Hispanic when I compare with non-Hispanic Whites.

In columns (5) and (6), I calculate the differences in means between HH-WW (column 5) and HW-WH (column 6). Overall, HH children do worse than WW children. HH has one year less of education compared to WW, and are 2 percentage points more likely to be unemployed. HH men earn 9\% less than WW men, and HH women make 2\% less than WW women. Therefore, these two groups, HH and WW, are different on many observable qualities.

The differentials between WH and HW are not as severe as the ones between HH and WW. HW children are still worse off than WH but are much better than HH. HW men attain 0.36 fewer years of education than WH men. HW women achieve 0.37 fewer years of schooling than WH women. HW children are 1 percentage points more likely to be unemployed. HW men earn 1\% less annual earnings as WH men, and HW women earn 3\% less. These differences show that HW and WH children are different from children of endogamous/homogamous marriages and, thus, are more comparable.

In table \ref{tab:c&p2}, I present the summary statistics for the four types and limit the sample to those that only identify as Hispanic.\footnote{Limiting the sample to those who identify as Hispanic only affects WH, HW, and HH. Members of the WW group are not affected by this restriction.} The same pattern discussed above between HH--WW and HW--WH persists. The one difference is that those who identify as Hispanic do slightly worse.

%table4
\input{tables/tab03a-summary-stats}

@@

Using the 1960 to 2000 censuses, I constructed a data set of synthetic parents. The sample includes married White men and women. Even though the census asks a person whether they are Hispanic or not, I took advantage of the questions on place of birth to create a proxy for ethnicity. I consider a Hispanic person as White and born in a Spanish-speaking country. Consequently, Whites in the sample are people who are White and native-born. Using the information provided in the census, I can link husbands and wives with each other. I assume that parents have children between the ages of 25 and 40. Therefore, my sample consists of married White men and women with children that were born in the 1920 to 1975 cohorts, see figure \ref{flowchart1} for a flowchart of the linking process.\footnote{The construction of "synthetic parents" follows the method used by \citet{rubinstein2014pride}.}

%table5
\input{tables/tab04-summary-stats}

I show the distribution of the four types of couples in table \ref{tab:mat2}. White husbands and White wives (WW) make up the majority of couples in the sample, 96\% (5,141,737 couples). Hispanic husbands and wives (HH) are the second largest group making up 2\% (119,749 couples) of all couples. White husbands and Hispanic wives (WH) couples are less 1\% (33,097 couples) of the sample and Hispanic husbands and White wives (HW) are also 1\% (37,847 couples).

I present the summary statistics of the four types of synthetic parents in table \ref{tab:synth}. WW couples have higher education, 12.75 years for husbands and 12.47 for wives. As a household, WW couples have 25.22 years of schooling. Additionally, the log total family income for a WW household is 10.72 and have 3.77 children. Husbands in HH marriages have 8.64 years of education, while women have 8.49 years of schooling. As a household, HH couples have 17.13 years of education, log total family income of 10.27, and have 4.23 children.

WH husbands have 11.77 years of education, while wives have an average of 10.40 years. WH household attained a total of 22.17 years of schooling in total, a log total family income of 10.55, and 3.98 children. HW husbands and wives obtained 10.25 and 11.11 years of education, respectively. HW household attained a total of 21.36 years of schooling in total, a log total family income of 10.46, and 4.15 children.

In columns 5 and 6 of table \ref{tab:synth}, I report the differences in means between HH-WW households and HW-WH households. An HH Hispanic husband attained 4.11 fewer years of education than a WW husband. While an HH wife acquired 3.98 fewer years of schooling than a White WW wife. The total educational difference between HH and WW households equals 8.09 years of education. The educational differences between HW and WH are less severe than those between HH and WW. A Hispanic husband in intermarriage attained 1.52 fewer years of schooling than a White husband in intermarriage. A White wife in intermarriage attained 0.70 years of education, more than a Hispanic wife in intermarriage. The difference in total years of education between HW and WH households is 0.81, which is an 90\% decrease from the HH--WW gap. The total family income of an HH household is 45\% lower than a WW household, while the total family income of an HW household is 10\% lower than a WH household. These differences show that couples that intermarry are better and different than those that do not.

@@

In this section, I present two empirical strategies. The first empirical strategy estimates the effect of having a Hispanic last name on annual log earnings. The second empirical strategy estimates the effect of identifying as Hispanic on log annual earnings.

@@

%table6
\input{tables/tab05-regression}
\subsection{Estimating the effect of having a Hispanic last name}

In this section I restrict the sample to WH and HW groups. Let $Y_{ist}$ be the annual log earnings of person $i$ in state $s$ at time $t$. $HW_{ist}$ is an indicator variables for the type of parents person $i$ has. $X_{ist}$ is a vector of controls that includes age and numbers of hours worked, $\gamma_{t}$ are year fixed effects, $\lambda_{s}$ are state fixed effects, and $\phi_{ist}$ represents the error term. The equation for this strategy is written as follows:
\begin{equation} \label{eq:1a}
Y_{ist} = \beta_{1} HW_{ist} + X_{ist} \pi + \gamma_{t} + \lambda_s + \phi_{ist}
\end{equation}

$\beta_{1}$ is the coefficient of interest in this specification. $\beta_{1}$ represents the earnings gaps between children of inter-ethnic marriages who have a Spanish-sounding last name versus a White last name.

\subsection{Estimating the effect of identifying as Hispanic}

In this section, I will present an estimation strategy that would allow me to capture the effect of identifying as a Hispanic.

\begin{equation} \label{eq:iden}
Y_{ist} = \beta_{1} HW_{ist} + \beta_{2} Hispanic_i + \beta_{3} HW_{ist} \cdot Hispanic_i + X_{ist} \pi + \gamma_{t} + \lambda_s + \phi_{ist}
\end{equation}

The coefficients of interest from Eq. \eqref{eq:iden} are $\beta_{1}$, $\beta_{2}$ and $\beta_{3}$. The coefficients $\beta_{1}$ and $\beta_{3}$ capture the effects of Spanish-sounding last names for $HW_{ist}$ that identify as Hispanic \textit{versus} those who do not.
%table7
\input{tables/tab06-regression}
\section{Results}\label{sec:results}

In this section, I present the results from estimating the two specifications in Eqs. \eqref{eq:1a} and \eqref{eq:iden}. The results are shown in Tables \ref{tab:lastnamereg} and \ref{tab:identreg}. All estimations are done on a sample of White, native-born men between the ages of 25 and 40 employed for full-time full-year (FTFY) and are wage and salary workers.

I find that an inter-ethnic person with a Spanish-sounding last names earns less than an inter-ethnic who has a White last name. A person with a Spanish-sounding last name earns 5 percentage points less than a person with a White last name. In other words, by comparing inter-ethnic children, a person with a Hispanic last name earns 5 percentage points less than someone with a White last name. However, more than half of the earnings gaps could be explained by educational differences. When I control for education, the last name effect decreases to a statistically insignificant 2 percentage points earnings gap. I also find a significant earnings gap between those that identify as Hispanic.

\subsection{The Effect of Having a Hispanic Last Name and Hispanic Identity on Labor Market Outcomes}

I provide the results to the estimation of Eq. \eqref{eq:1a} in table \ref{tab:lastnamereg}. The sample in this analysis includes full-time, full-year, and wage and salary men, and I control for hours worked and age and include year fixed effects (FE). I also restrict the sample to children of HW and WH (omitted) parents. Column one in table \ref{tab:lastnamereg} is the results without controlling for education. Overall, an inter-ethnic with a Hispanic last name earns 5 percentage points less than an inter-ethnic with a White last name. This gap, however, could be entirely explained by educational differences. An inter-ethnic with a Hispanic last name earns 2 percentage points less than an inter-ethnic with a White last name, but the result is statistically insignificant.

I provide the results to the estimation of Eq. \eqref{eq:iden} in table \ref{tab:identreg}. The sample in this analysis includes full-time, full-year, and wage and salary men, and I control for hours worked and age and include year fixed effects (FE). I also restrict the sample to children of HW and WH parents. The omitted group is children of WH parents that do not identify as Hispanic. Column one in table \ref{tab:identreg} is the results without controlling for education. An inter-ethnic $HW_{ist}$ that identifies as Hispanic \textit{versus} $HW_{ist}$ who do not---i.e. $HW_{ist} + HW_{ist} \cdot Hispanic_i$---earns 5 percentage points less than an inter-ethnic that does not, but the result could be entirely explained by educational differences.

@@

As the Hispanic population grows in the United States, studying discrimination against this group becomes increasingly important. In this paper, I examine discrimination against Hispanics in the labor market. More specifically, I examine the impact of Hispanic last names and Hispanic identification on annual log earnings.

I compare the children of inter-ethnic marriages to study the labor market impact of having a Hispanic last name. When I compare the earnings of HW and WH children, which captures the effect of having a Hispanic last name, HW children earn  5 percentage points less than WH children. Thus, by comparing inter-ethnic children, a person with a Hispanic last name makes 5 percentage points less than someone with a White last name. When I control for education, the last name effect decreases to a statistically insignificant 2 percentage points earnings gap. Moreover, I estimate the effect of identifying as Hispanic and having a Hispanic last name. I find that an inter-ethnic child with a Hispanic last name and identifies as Hispanic earns 5 percentage points less that an inter-ethnic child with native sounding last name and identifies as Hispanic. The earnings gap, however, could be entirely explained by educational differences.

While the earnings gap between children of inter-ethnic parents with and without a Hispanic last name disappears when controlling for education, it does not necessarily indicate the absence of discrimination. Education itself can be influenced by bias, potentially resulting in divergent outcomes. Consequently, further research is essential to comprehensively understand the earnings gaps between Hispanics and Whites.

@@

\bibliographystyle{cas-model2-names}
\bibliography{bibliography}

>>>>

\affiliation[]{organization={Department of Economics, Tulane University},
            addressline={6823 St. Charles Ave.},
            city={New Orleans},
            % citysep={}, % Uncomment if no comma needed between city and postcode
            % postcode={70118},
            state={LA 70118},
            country={United States}}

\begin{abstract}
\singlespacing % Set single spacing for the abstract
Do individuals with Hispanic names face labor market discrimination? This study analyzes the impact of Hispanic-sounding surnames on wages among inter-ethnic children with one White and one Hispanic parent. I find that individuals with Hispanic surnames often earn less, with a notable wage gap favoring those with White surnames. People born to Hispanic fathers and White mothers receive 0.3 years of education less than those born to White fathers and Hispanic mothers. Males born to Hispanic fathers and White mothers earn 5 percentage points less than those born to White fathers and Hispanic mothers, largely due to educational differences. Additionally, men with Spanish-sounding last names who identify as Hispanic earn significantly less, again due to educational disparities.
\end{abstract}

@@

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Main Text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

A large literature documents substantial earnings gaps across race and ethnicity \citep{bayer2018divergent, charles2008prejudice, card1992school, fryer2004causes, rubinstein2014pride, bertrand2004emily, juhn1991accounting}. Hispanics constitute a large and growing portion of the population in the United States. As the number of Hispanics increases, determining whether ethnic discrimination affects their labor market outcomes becomes increasingly crucial \citep{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017,abramitzkyImmigrantsAssimilateMore2020a, abramitzkyNationImmigrantsAssimilation2014,abramitzkyCulturalAssimilationAge2016,chettyWhereLandOpportunity2014}. Thus, it is important to understand whether a person's ethnicity affects their labor market outcomes. Assimilation and mobility are crucial because they reflect how well Hispanics can integrate into society and move up the socioeconomic ladder.

In this paper, I answer the following question. Does having a Hispanic last name affect education and labor market outcomes? Moreover, I aim to show that comparing Hispanic Whites to non-Hispanic Whites might create an artificially higher earnings gap since the two groups differ on many observable characteristics. \footnote{Observable characteristics refer to factors that can be measured and quantified, such as education level, work experience, and immigration status.} Others have attempted to compare how native-born White Hispanics fare to non-Hispanic Whites and foreign-born Hispanics. In \citet{antman2020ethnic,antmanEthnicAttritionObserved2016,antmanEthnicAttritionObserved2016a,antmanEthnicAttritionAssimilation2020}, the authors compare the health and educational outcomes of Hispanic Whites to non-Hispanic Whites and native-born Hispanics to foreign-born Hispanics. They find gaps in education and health between Hispanics and Whites. They also find that native-born Hispanics are more likely than their foreign-born counterparts to report poor health. \citet{davilaChangesRelativeEarnings2008} documents many gaps in labor market outcomes between Hispanics and Whites. They attribute a big part of this gap to differences in education, experience, immigration status, and regional differences.

The US population is growing in diversity. The proportion of non-Whites has increased by more than 10 percentage points from 13 percent in 1995 to 23 percent in 2019. The number of Hispanics has grown by 9 percentage points from 9 percent in 1995 to 16 percent in 2019.\footnote{The portion of non-Whites and Hispanics is calculated using the Current Population Survey (CPS).} Native-born White Hispanic men earn 21\% less than White men, although a substantial portion of the earnings gap is due to educational differences between Hispanics and Whites \citep{duncan2006hispanics, duncan2018identifying, duncan2018socioeconomic}. Some of the earnings differences may also be due to discrimination which will have negative consequences. For example, discrimination against Hispanics can lead to reduced job opportunities, lower wages, and hinder assimilation. In this paper, I examine the role of having a Hispanic last name on labor market and educational outcomes outcomes.

Identifying discrimination is difficult because of factors that affect labor market outcomes that are unobservable to economists---such as unobserved skills and separating it from prejudice and stereotypes. One strategy used by researchers is audit or resume studies. \citet{bertrand2004emily} conducted an audit study where identical resumes were sent to employers with White and Black-sounding names. This approach, however, has its drawbacks. Audit studies only observe callbacks, not wages.

This study utilizes a method developed by \citet{rubinstein2014pride}. I compare children from inter-ethnic marriages. More precisely, I compare children of Hispanic fathers and White mothers (henceforth HW) to children of White fathers and Hispanic mothers (henceforth  WH). This approach stems from the fact that there is a strong selection among many characteristics, and thus, marriages are not random. Couples match on several observable characteristics like income, schooling, socio-economic background, etc. \citep{averettBetterWorseRelationship2008, averettEconomicRealityBeauty1996, beckerTheoryMarriagePart1973, beckerTheoryMarriagePart1974, beckerTreatiseFamily1993, browningCollectiveUnitaryModels2006, chiapporiFatterAttractionAnthropometric2012}. Children of HW and WH marriages have more similar observable characteristics than children of endogamous/homogamous marriages--- i.e., White fathers-White mothers and Hispanic fathers-Hispanic mothers. Moreover, children from a Hispanic father and White mother household will have a Hispanic last name from their fathers, enabling the investigation of how ethnic signals, such as having a Hispanic last name, affect annual log earnings.

The main identifying assumption of my empirical strategy depends on the assumption that people born to HW parents are similar to their WH peer in all observable---and unobservable---aspects and characteristics that are important in the labor market. Consequently, the only difference between the two groups is the variation in which group is more likely to have a Hispanic sounding last name. Children from mixed ethnic backgrounds may appear physically similar to those from single ethnicity backgrounds, but the influence of family dynamics and upbringing, crucial in developing skills and personal characteristics, varies with the pattern of mixed ethnic marriages. Particularly in a society where Hispanic ancestry is perceived negatively, it raises the question of what kind of White women would choose Hispanic men. Furthermore, even if such unions were formed randomly, children from White-Hispanic homes might benefit from more favorable family conditions than those from Hispanic-White homes, considering Whites generally have stronger socio-economic backgrounds than Hispanics. These factors introduce doubts regarding whether children from Hispanic-White families receive comparable familial support and influences, either genetically or environmentally, as those from White-Hispanic families.

@@

I use two datasets: the IPUMS Current Population Survey (CPS) Annual Social and Economic (ASEC) \citep{cps2019} and the 1960 to 2000 US censuses \citep{acs2019}.

The CPS data (1994-2019) is used to study the effect of Hispanic last names on labor market outcomes. The census data is employed to construct synthetic parents using the method developed by \citet{rubinstein2014pride}.

@@

The CPS sample is restricted to White, US-born citizens aged 25-40, born between 1960-2000. Parents are classified as Hispanic if born in a Spanish-speaking country or Puerto Rico, and White if US-born. Four parental types are identified: (1) White father and White mother (WW), (2) White father and Hispanic mother (WH), (3) Hispanic father and White mother (HW), and (4) Hispanic father and Hispanic mother (HH).

Table \ref{tab:combined-mat} shows the distribution of these types. WW children comprise 96\% of the sample, HH 3\%, and inter-ethnic children (WH and HW) 1.35\%. Summary statistics for each group are presented in Table \ref{tab:combined-sum}.

@@

Using the 1960-2000 censuses, I constructed a dataset of synthetic parents, including married White men and women born between 1920-1975. Hispanic individuals are defined as White and born in a Spanish-speaking country, while Whites are native-born.

Table \ref{tab:combined-mat} shows the distribution of couple types: WW (96\%), HH (2\%), WH (<1\%), and HW (1\%). Summary statistics for the parents are presented in Table \ref{tab:combined-sum}.

@@

In this section, I present two empirical strategies. The first empirical strategy estimates the effect of having a Hispanic last name on educational outcomes and earnings.

@@

\subsection{Estimating the effect of having a Hispanic last name}

In this section I restrict the sample to WH and HW groups. Let $Y_{ist}$ be the outcome of interest for person $i$ in state $s$ at time $t$. $HW_{ist}$ is an indicator variables for the type of parents of person $i$ has. $X_{ist}$ is a vector of controls that includes age and numbers of hours worked, $\gamma_{st}$ are state-year fixed effects (FE), and $\phi_{ist}$ represents the error term. The equation for this strategy is written as follows:
\begin{equation} \label{eq:1a}
Y_{ist} = \beta_{1} HW_{ist} + X_{ist} \pi + \gamma_{st} + \phi_{ist}
\end{equation}

$\beta_{1}$ is the coefficient of interest in this specification. $\beta_{1}$ represents the gaps in outcomes between children of inter-ethnic marriages who have a Spanish-sounding last name versus a White last name.

\subsection{Threats to Identification}

The central assumption of my estimation strategy is that individuals born to HW parents have comparable characteristics to their WH peers, particularly in areas crucial to the labor market. This assumption faces two main challenges. First, Selection in the marriage market: Given the potential societal penalties associated with Hispanic identity, the characteristics of White women who marry Hispanic men may be unique. Second, Differential parental influence: Fathers and mothers may impact human capital accumulation differently \citep{kimball2009risk,magruder2010intergenerational}.

These concerns are mitigated by assortative matching in marriages. A substantial body of literature indicates strong selection in marriages based on traits \citep{averettBetterWorseRelationship2008, beckerTreatiseFamily1993,duncanIntermarriageIntergenerationalTransmission2011}. This suggests that inter-ethnic WH and HW marriages likely consist of partners with similar characteristics, supporting the comparability of their children in terms of labor market outcomes.

A potential measurement error arises from using parents' birthplace in the CPS data as a proxy for ethnicity and last name. However, this concern is minimal as most Hispanics from 1960 to 2000 were first-generation immigrants. Census data shows only 3\% of native-born Americans identified as Hispanic, making it highly improbable for an inter-ethnic child to have a native-born, second-generation+ Hispanic father.\footnote{Self-reported Hispanic identity among first-generation Hispanic immigrants and native-borns are based on the author’s calculations from the 1960-2000 Census. The sample includes Whites, who are married, and are between the ages 25 and 40.} Empirical evidence from the CPS and US Census will be used to evaluate these concerns and support the validity of the identification strategy.

\section{From the Data: The Differences Between HW and WH Couples}\label{sec:hw-wh-couples-data}

In this section, I examine the data to validate my empirical strategy, focusing on educational and economic outcomes for four ethnic groupings—White White (WW), White Hispanic (WH), Hispanic White (HW), and Hispanic Hispanic (HH)—as shown in Table \ref{tab:combined-sum}. The data indicates marriage selection, with smaller differences between HW and WH children compared to the other groups, making them a more suitable comparison for analyzing labor market discrimination against Hispanics. Using the synthetic parents sample constructed from Census data, we observe that WW couples have the highest household education levels (24.95 years), while HH couples have the lowest (17.13 years). Inter-ethnic couples fall in between, with WH households totaling 22.68 years of schooling and HW households slightly behind at 21.50 years. Notably, HW wives are more educated than their WH peers, which could have implications for the educational attainment of HW children, as mothers play a crucial role in human capital accumulation.

Regarding labor market outcomes, WW households have the highest log total family income (10.75), followed by WH (10.65), HW (10.60), and HH (10.42) households. The earnings gap between HW and WH men is marginal, with HW men earning 4\% less. However, HW women slightly out-earn WH women, with log hourly earnings of 1.75 and 1.73, respectively. This reversal in typical gender earning patterns suggests that HW women contribute more to their families' economic stability, potentially benefiting their children's development and future labor market prospects. Despite HW mothers' higher education and income levels, HW children complete, on average, 0.4 fewer years of education than their WH peers (Table \ref{tab:combined-sum}), potentially pointing to discrimination or barriers in educational access for HW children.

\section{Results}\label{sec:results}

\subsection{Effect of Hispanic Last Name on Educational Outcomes}

Analysis of White, U.S.-born Hispanics ages 25-40 (Table \ref{tab:lastname-ed-reg}) reveals that HW children receive 0.2 years less education than WH children. While there is no significant difference in high school dropout rates, HW children are 2 percentage points less likely to earn an associate degree and 3 percentage points less likely to earn a bachelor's degree. These gaps are larger for HW women in higher education. These disparities suggest potential barriers or discrimination in access to higher education for HW children, particularly women.

\subsection{Effect on Labor Market Outcomes}

Results from Table \ref{tab:lastnamereg} indicate a 1 percentage point employment gap between HW and WH workers, and a 5 percentage point crude earnings gap. However, when controlling for education, the earnings gap reduces to a statistically insignificant 1 percentage point.

@@

This study examines discrimination against Hispanics in the U.S. labor market, focusing on the impact of Hispanic last names and identification on education and earnings. The research finds that inter-ethnic individuals with Spanish-sounding last names receive 0.2 years less education and are more likely to be unemployed. HW children earn 5 percentage points less than WH children, but this gap becomes statistically insignificant when controlling for education.

These results align with existing literature on disparities in educational access \citep{bergman2018education,gaddis2024racial}. While the earnings gap diminishes when controlling for education, this doesn't necessarily indicate an absence of discrimination, as education itself can be influenced by bias.

Further research is needed to comprehensively understand the earnings gaps between Hispanics and Whites, particularly considering that mothers of HW children have higher education and earnings, which theoretically should lead to better outcomes for their children.

@@

\newpage
\clearpage

\bibliographystyle{cas-model2-names}
\bibliography{bibliography}
\newpage
\clearpage
%table1
\input{tables/tab16-tab1-and-tab2}

\newpage
\clearpage

%table3
\input{tables/tab17-all-sums}
